---
title: "Short Introduction to BAMP"
author: "Volker Schmid"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Short Introduction to BAMP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library(bamp)
```

## Introduction

Age-period-cohort (APC) models decompose incidence or mortality rates into
three smooth components: an **age** effect (how risk varies across age
groups), a **period** effect (how risk varies over calendar time) and a
**cohort** effect (how risk varies across birth cohorts). BAMP fits such
models in a Bayesian framework via MCMC and provides tools to check
convergence, inspect the fitted effects, and predict future cases.

## Data example

BAMP includes a data example: `cases` and `population`, matrices of observed
case counts and population at risk, cross-classified by calendar year (rows)
and age group (columns).

```{r loadplot}
data(apc)
plot(cases[,1],type="l",ylim=range(cases), ylab="cases", xlab="year", main="cases per age group")
for (i in 2:8)lines(cases[,i], col=i)
```

For simulating your own APC data, see *vignette("simulation", package="bamp")*.

## APC model with random-walk first-order priors

We fit a model with a first-order random walk (`"rw1"`) prior on each of the
age, period and cohort effects -- a common default that assumes each effect
changes smoothly, without a systematic trend, between neighbouring levels.
`periods_per_agegroup` tells `bamp()` how many calendar years each age group
spans, which is needed to map age and period onto the cohort dimension.

```{r Model1_with_RW1_priors}
model1 <- bamp(cases, population, age="rw1", period="rw1", cohort="rw1",
              periods_per_agegroup = 5)
```

`bamp()` automatically checks MCMC convergence using the Gelman-Rubin
diagnostic and warns if the chains have not converged. We can also run this
check manually:

```{r check_model}
checkConvergence(model1)
```

`print()` summarises the fitted model: posterior estimates of the smoothing
(precision) parameters, and the deviance and DIC, which are useful for
comparing models (see *vignette("modeling", package="bamp")* for model
selection):

```{r print_model}
print(model1)
```

The fitted age, period and cohort effects can be plotted with point-wise
posterior quantiles:

```{r plot_model}
plot(model1)
```

By default the plot shows the median and a 90% interval; other quantiles can
be requested explicitly:

```{r plot_model_with_more_quantiles}
plot(model1, quantiles = c(0.025,0.1,0.5,0.9,0.975))
```

For other prior choices (e.g. second-order random walks, heterogeneity,
covariates), see *vignette("modeling", package="bamp")*.

## Prediction

Because the period and cohort effects are modelled as random walks, their
most likely continuation is a flat extrapolation with growing uncertainty --
which lets us predict cases for upcoming years:

```{r prediction}
pred <- predict_apc(object=model1, periods=3)
```

The plot below shows the predicted probability of a case per age group; the
dashed vertical line marks the start of the 3-year forecast:

```{r}
m<-max(pred$pr[2,,])
plot(pred$pr[2,,8],type="l", ylab="probability", xlab="year", ylim=c(0,m))
for (i in 7:1)
  lines(pred$pr[2,,i],col=8-i)
legend(1,m,col=8:1,legend=paste("Age group",1:8),lwd=2,cex=0.6)
lines(c(10.5,10.5),c(0,1),lty=2)
```

For more details, see *vignette("prediction", package="bamp")*.
