---
title: "Bayesian Age-Period-Cohort Modeling"
author: "Volker Schmid"
date: "`r Sys.Date()`"
#output: rmarkdown::html_vignette
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Bayesian Age-Period-Cohort Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library(bamp)
```

This vignette walks through the range of age-period-cohort (APC) models that
`bamp()` can fit: different smoothness priors for the age, period and cohort
effects, a model that omits an effect entirely, and models where the period
or cohort effect is scaled by a known covariate. For a first introduction to
`bamp()`, see *vignette("bamp", package="bamp")*; for forecasting, see
*vignette("prediction", package="bamp")*.

## Data example

BAMP includes a data example: `cases` and `population`, matrices of observed
case counts and population at risk, cross-classified by calendar year (rows)
and age group (columns). It also bundles `cov_p` and `cov_c`, example period
and cohort covariates used in the last two models below.

```{r loadplot}
data(apc)
plot(cases[,1],type="l",ylim=range(cases), ylab="cases", xlab="year", main="cases per age group")
for (i in 2:8)lines(cases[,i], col=i)
```

## APC model with random-walk first-order priors

A first-order random walk (`"rw1"`) prior assumes each effect changes
smoothly, without a systematic trend, between neighbouring age groups,
periods or cohorts. This is the most common default.

```{r Model1_with_RW1_priors}
model1 <- bamp(cases, population, age="rw1", period="rw1", cohort="rw1",
              periods_per_agegroup = 5)
```

`bamp()` automatically checks MCMC convergence using the Gelman-Rubin
diagnostic. We can also run this check manually:

```{r check_model}
checkConvergence(model1)
```

`print()` summarises the fitted model: posterior estimates of the smoothing
(precision) parameters, and the deviance and DIC, which are useful for
comparing the models fitted in this vignette:

```{r print_model}
print(model1)
```

The fitted age, period and cohort effects can be plotted with point-wise
posterior quantiles:

```{r plot_model}
plot(model1)
```

By default the plot shows the median and a 90% interval; other quantiles can
be requested explicitly:

```{r plot_model_with_more_quantiles}
plot(model1, quantiles = c(0.025,0.1,0.5,0.9,0.975))
```

## APC model with random-walk second-order priors

A second-order random walk (`"rw2"`) prior penalises curvature rather than
level, so it produces effects that follow a smooth, roughly linear trend
instead of reverting towards a constant. RW2 priors are less informative
than RW1, so the chains mix more slowly and need substantially more
iterations to converge; here we also set the smoothing hyperparameters
explicitly instead of using the defaults.

```{r Model2_with_RW2_priors}
model2 <- bamp(cases, population, age="rw2", period="rw2", cohort="rw2",
              periods_per_agegroup = 5,
              mcmc.options=list("number_of_iterations"=200000, "burn_in"=100000, "step"=50, "tuning"=500),
              hyperpar=list("age"=c(1,.5), "period"=c(1,0.05), "cohort"=c(1,0.05)))
```

```{r Model2_results}
checkConvergence(model2)
print(model2)
plot(model2)
```

## APC model without a period effect

Age, period and cohort are linearly dependent -- any one is determined by
the other two -- so a model with all three as unrestricted smooth trends is
not identifiable. One common way to sidestep this is to drop the effect of
least interest -- here the period effect (`period=" "`) -- and keep only age
and cohort:

```{r model3}
model3<-bamp(cases, population, age="rw1", period=" ", cohort="rw2",
              periods_per_agegroup = 5)
checkConvergence(model3)
print(model3)
plot(model3)
```

## APC model with a cohort covariate

Instead of dropping an effect, a known covariate can scale it: `cohort_covariate`
multiplies the cohort effect by a given value at each cohort (here the bundled
example covariate `cov_c`), which is useful when an external, cohort-specific
risk factor is available.

```{r model4}
(model4<-bamp(cases, population, age="rw1", period="rw1", cohort="rw1",
             cohort_covariate = cov_c, periods_per_agegroup = 5))
plot(model4)
```

## APC model with a period covariate

Analogously, `period_covariate` scales the period effect by a known,
period-specific covariate (here `cov_p`):

```{r model5}
(model5<-bamp(cases, population, age="rw1", period="rw1", cohort="rw1",
             period_covariate = cov_p, periods_per_agegroup = 5))
plot(model5)
```
