---
title: "Simulating Age-Period-Cohort Data"
author: "Volker Schmid"
date: "`r Sys.Date()`"
#output: rmarkdown::html_vignette
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Simulation of Age-Period-Cohort Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, out.width = "90%")
library(bamp)
```

This vignette simulates age-period-cohort data from known, hand-specified
age, period and cohort effects, fits `bamp()` to the simulated cases, and
checks that the fitted effects recover the true ones -- a good way to sanity
check the method, or to plan how much data a study design would need. For a
first introduction to `bamp()`, see *vignette("bamp", package="bamp")*; for
other model specifications, see *vignette("modeling", package="bamp")*.

## Specifying the true effects

We first fix the three true effects on the log-odds scale; each is centred
(mean zero) since only their shape, not their level, is identified (see
*vignette("modeling", package="bamp")*). The age effect is an increasing,
concave curve:

```{r simulate_age}
age=2*sqrt(seq(1,20,length=10))
age<- age-mean(age)
plot(age, type="l")
```

The period effect is V-shaped: it falls for the first half of the observed
periods, then rises again over the second half:

```{r simulate_period}
period=25:1
period[13:25]<-13:25
period<-period/5
period<-period-mean(period)
plot(period, type="l")
```

The cohort effect falls for the earliest cohorts, rises through the middle
ones, then slightly falls for the most recent cohorts. With
`periods_per_agegroup = 5` and 10 age groups spanning 25 periods, there are
`periods_per_agegroup * (10 - 1) + 25 = 70` cohorts:

```{r simulate_cohort}
periods_per_agegroup=5
number_of_cohorts <- periods_per_agegroup*(10-1)+25
cohort<-rep(0,70)
cohort[1:20]<-(19:0)
cohort[21:40]<- (1:20)/2
cohort[41:70]<- 10
cohort=cohort-(1:70)/10
cohort<-cohort/10
cohort<-cohort-mean(cohort)
plot(cohort, type="l")
```

## Simulating data from the true effects

`apcSimulate()` combines an intercept (here `-10`, on the log-odds scale)
with the three true effects and a population of `1 000 000` per Lexis cell to draw
simulated case counts, returning `cases` and `population` matrices just like
the bundled `apc` data example:

```{r simulate_data}
simdata<-apcSimulate(-10, age, period, cohort, periods_per_agegroup, 1e6)
print(simdata$cases)
```

## Recovering the effects with `bamp()`

We fit the same first-order random-walk model used throughout the other
vignettes to the simulated data:

```{r bamp}
simmod <- bamp(cases = simdata$cases, population = simdata$population, age = "rw2",
period = "rw2", cohort = "rw2", periods_per_agegroup =periods_per_agegroup)
```

```{r check_print_plot}
print(simmod)
par(mfrow=c(3,1))
plot(simmod)
```

Since we know the true effects here (unlike with real data), we can check
directly how well `bamp()` recovered them: the black line is the true
effect, the blue line the fitted posterior median (`effects()`). Because the
age, period and cohort effects are only identified up to a shared linear
trend (see *vignette("modeling", package="bamp")*), the recovered curves
show the same nonlinear (curvature) features as the truth, but are
linearly shifted relative to it.

```{r plot_comparison}
effects<-effects(simmod)
plot(age, type="l")
lines(effects$age, col="blue")
plot(period, type="l", ylim=range(effects$period))
lines(effects$period, col="blue")
plot(cohort, type="l", ylim=range(effects$cohort))
lines(effects$cohort, col="blue")
```

## Prediction

As in *vignette("prediction", package="bamp")*, the fitted random walks can
be extrapolated to forecast cases for periods beyond the simulated data --
here 5 more, so the supplied `population` array covers 30 periods (25
observed + 5 forecast) across the 10 age groups:

```{r prediction}
prediction<-predict_apc(simmod, periods=5, population=array(1e6,c(30,10)), quantiles=c(.1,.5,.9))
```

Total predicted cases per period are shown with their median
(solid points) and 80% credible interval (dashes):

```{r prediction1}
plot(prediction$cases_period[2,], ylim=range(prediction$cases_period),ylab="",pch=19)
points(prediction$cases_period[1,],pch="–",cex=2)
points(prediction$cases_period[3,],pch="–",cex=2)
for (i in 1:30)lines(rep(i,3),prediction$cases_period[,i])
```

The period effect itself extends the same way into the 5 forecast periods:

```{r prediction4}
plot(prediction$period[2,], type="l", ylim=range(prediction$period))
points(prediction$period[2,1:25])
for (i in 26:30)lines(rep(i,3),prediction$period[,i])
```

## Period covariate

Finally, we refit the simulated data with a period covariate scaling the
period effect (see *vignette("modeling", package="bamp")*). A period
covariate is a positive multiplier, so it must be larger than zero.

```{r simulate_covariate_for_period}
cov_p<-exp(rnorm(30,period,.1))
```

```{r bamp_with_covariate}
simmod2 <- bamp(cases = simdata$cases, population = simdata$population, age = "rw1",
period = "rw1", cohort = "rw1", periods_per_agegroup = periods_per_agegroup,
period_covariate = cov_p)
```

```{r check_print_plot2}
print(simmod2)
par(mfrow=c(3,1))
plot(simmod2)
```

```{r prediction2}
prediction2<-predict_apc(simmod2, periods=5, population=array(1e6,c(30,10)), quantiles=c(.1,.5,.9))
```

```{r prediction3}
plot(prediction2$cases_period[2,], ylim=range(prediction2$cases_period),ylab="",pch=19)
points(prediction2$cases_period[1,],pch="–",cex=2)
points(prediction2$cases_period[3,],pch="–",cex=2)
for (i in 1:30)lines(rep(i,3),prediction2$cases_period[,i])
```