---
title: "Bayesian Age-Period-Cohort Modelling"
author: "Volker Schmid"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Age-Period-Cohort Modelling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library(bamp)
```

## Data example

BAMP includes a data example.

```{r load_and_plot_example_data_set}
data(apc)
plot(cases[,1],type="l",ylim=range(cases), ylab="cases", xlab="year", main="cases per age group")
for (i in 2:8)lines(cases[,i], col=i)
```

## APC model with random walk first order prior
```{r Model1_with_RW1_priors}
model1 <- bamp(cases, population, age="rw1", period="rw1", cohort="rw1",
              periods_per_agegroup = 5)
```

bamp() automatically performs a check for MCMC convergence using Gelman and Rubin's convergence diagnostic. We can manually check the convergence again:

```{r check_model}
checkConvergence(model1)
```

Now we have a look at the model results. This includes estimates of smoothing parameters and deviance and DIC:
```{r print_model}
print(model1)
```

We can plot the main APC effects using point-wise quantiles: 
```{r plot_model}
plot(model1)
```

More quantiles are possible:
```{r plot_model_with_more_quantiles}
plot(model1, quantiles = c(0.025,0.1,0.5,0.9,0.975))
```

## APC model with random walk first order prior
```{r Model2_with_RW2_priors}
model2 <- bamp(cases, population, age="rw2", period="rw2", cohort="rw2",
              periods_per_agegroup = 5)
```

```{r Model2_results}
checkConvergence(model2)
print(model2)
plot(model2)
```

```{r model3}
model3<-bamp(cases, population, age="rw1", period=" ", cohort="rw2",
              periods_per_agegroup = 5)
checkConvergence(model3)
print(model3)
plot(model3)
```

```{r model4}
(model4<-bamp(cases, population, age="rw1", period="rw1", cohort="rw1",
             cohort_covariate = cov_c, periods_per_agegroup = 5, 
             parallel = FALSE, 
             mcmc.options=list("number_of_iterations"=50000)
             ))
plot(model4)
```


```{r model5}
(model5<-bamp(cases, population, age="rw1", period="rw1", cohort="rw1",
             period_covariate = cov_p, periods_per_agegroup = 5, 
             parallel = FALSE))
plot(model5)
```
