---
title: "Bayesian Age-Period-Cohort Modelling"
author: "Volker Schmid"
date: "16th November 2017"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Bayesian Age-Period-Cohort Modelling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library(bamp)
```

## Data example

BAMP includes a data example.

```{r load and plot example data set}
data(apc)
plot(cases[,1],type="l",ylim=range(cases), ylab="cases", xlab="year", main="cases per age group")
for (i in 2:8)lines(cases[,i], col=i)
```

## APC model with random walk first order prior
```{r Model 1 with RW1 priors}
model1 <- bamp(cases, population, age="rw1", period="rw1", cohort="rw1",
              periods_per_agegroup = 5,  verbose=TRUE)
```

bamp() automatically performs a check for MCMC convergence using Gelman and Rubin's convergence diagnostic. We can manually check the convergence again:

```{r check model}
check(model1)
```

Now we have a look at the model results. This includes estimates of smoothing parameters and deviance and DIC:
```{r print model}
print(model1)
```

We can plot the main APC effects using point-wise quantiles: 
```{r plot model}
plot(model1)
```

More quantiles are possible:
```{r plot model with more quantiles}
plot(model1, quantiles = c(0.025,0.1,0.5,0.9,0.975))
```
