---
title: "Short Introduction to BAMP"
author: "Volker Schmid"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Bayesian Age-Period-Cohort Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library(bamp)
```

## Data example

BAMP includes a data example.

```{r loadplot}
data(apc)
plot(cases[,1],type="l",ylim=range(cases), ylab="cases", xlab="year", main="cases per age group")
for (i in 2:8)lines(cases[,i], col=i)
```

For simulating APC data, see *vignette("simulation", package="bamp")*.

## APC model with random walk first order prior
```{r Model1_with_RW1_priors}
model1 <- bamp(cases, population, age="rw1", period="rw1", cohort="rw1",
              periods_per_agegroup = 5)
```

bamp() automatically performs a check for MCMC convergence using Gelman and Rubin's convergence diagnostic. We can manually check the convergence again:

```{r check_model}
checkConvergence(model1)
```

Now we have a look at the model results. This includes estimates of smoothing parameters and deviance and DIC:
```{r print_model}
print(model1)
```

We can plot the main APC effects using point-wise quantiles: 
```{r plot_model}
plot(model1)
```

More quantiles are possible:
```{r plot_model_with_more_quantiles}
plot(model1, quantiles = c(0.025,0.1,0.5,0.9,0.975))
```

For other models see *vignette("modeling",package="bamp")*.

## Prediction

Using the prior assumption of a random walk for the period and cohort effect, one can predict cases for upcoming years. 

```{r prediction}
pred <- predict_apc(object=model1, periods=3)
```

```{r}
m<-max(pred$pr[2,,])
plot(pred$pr[2,,8],type="l", ylab="probability", xlab="year", ylim=c(0,m))
for (i in 7:1)
  lines(pred$pr[2,,i],col=8-i)
legend(1,m,col=8:1,legend=paste("Age group",1:8),lwd=2,cex=0.6)
lines(c(10.5,10.5),c(0,1),lty=2)
```

More details see *vignette("prediction",package="bamp")*.
